---
title: "Exponential Simulation and Inference"
author: "Alex Fennell"
date: "2/11/2022"
output: pdf_document
---

# Simulating the Exponential distribution

This project will compare how statistics of a non-normal distribution are 
themselves normally distributed. A Monte Carlo procedure will be used to simulate
distributions of the mean and standard deviation of the exponential distribution
which will then be compared to the theoretical mean and standard deviation of the
distribution. 

```{r}
library(ggplot2)

```

## Simulations
This simulation will use samples of size 40 with lambda=.2. The theoretical mean
and standarddeviation can be calculated as 1/lambda. 

``` {r}
lambda=.2
#Theoretical mean and SD
Tmean<-1/lambda
Tsd<-1/lambda
#Number of Monte Carlo samples
S=1000
#vectors to store means and SDs of Monte Carlo samples
means<-vector(mode="numeric",length=S)
sds<-vector(mode="numeric",length=S)
for (i in 1:S){
#Sample sizes of 40 drawn from the exponential distribution with rate lambda
    samp<-rexp(40,lambda)
    means[i]<-mean(samp)
    sds[i]<-sd(samp)
}
```

## Sample Mean vs. Theoretical Mean
This next chunk of code will show a plot of the means from the Monte Carlo 
simulation with the theoretical mean overlayed as a vertical line.
``` {r,warnings=False,message=FALSE}
meanplot<-ggplot()+geom_histogram(aes(means),color="black",fill="lightblue")
meanplot+geom_vline(xintercept=Tmean,linetype="dashed",color="red",size=1.5) +
    ggtitle("Distribution of Means of Exponential Simulation") +
    xlab("Means")+
    theme(plot.title = element_text(hjust = 0.5))
    
```
