---
title: "Inferential Analysis of Tooth Decay"
author: "Alex Fennell"

output: pdf_document
---
# Synopsis

The purpose of this analysis is to examine the Tooth Growth data set and perform
some basic inferential statistics on it. The first step of this analysis is an
exploratory data analysis, which will inform how to go about an inferential analysis.


```{r libraries}
library(dplyr)
library(ggplot2)
```


## Reading in the data
The Tooth Growth data set is already available as it is part of the R datasets
package. Using the str function, we can see what the basic structure of the dataset 
is.

```{r tooth data exploration}
TGdata<-ToothGrowth
str(TGdata)
```

This shows the dataset with 3 variables, len-length of tooth growth, 
supp-The supplement type, and dose-the numeric does in mg per day. The first thing
we can do is change dose to a factor variable and then look at various summary statistics
as a function of the two factor variables, dose and supplement type.

```{r initial data summary,message=FALSE,warning=FALSE}   
# Need to change dose column into a factor rather than a numeric variable
TGdata$dose<-as.factor(TGdata$dose)

Summary_TGdata<-TGdata %>%
    group_by(supp,dose)%>%
    summarise(MEAN=mean(len),SD=sd(len),MAX=max(len),MIN=min(len),COUNT=n())
Summary_TGdata
```

This output shows that mean length of odontoblasts increases with dose across both
supplement types. The maximum and minimum values also increase as a funciton of 
increasing dosage. The sample size is the same across all supplement type and
dose manipulations. It is difficult to tell if the mean odontoblast length differs
across supplement types, so in the next block of code, we will examine this.

```{r tooth data mean comparison}
suppcomp<-TGdata %>%
    group_by(supp)%>%
    summarise(MEAN=mean(len),SD=sd(len),MAX=max(len),MIN=min(len))
suppcomp
```
 
 This shows that on average the mean odntoblast length for guinea pigs given 
 orange juice is longer than guinea pigs given ascorbic acid. In addition to this,
 the maximum and minimum values for guinea pigs in the OJ group are not as extreme
 as those in the VC group. 
 
 ### Boxplot analysis

